---
title: "Calgary Crime Analysis"
author: "Marc Boulet"
date: "9/27/2017"
output: 
  html_document: 
    theme: spacelab
---

```{r setup, include=FALSE, message=FALSE, error=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calgary Crime Statistics

We will perform a data analysis on the following dataset from the following website: http://www.calgary.ca/cps/Documents/statistical-reports/2017%20Community%20Crime%20Statistics.xls.


The column names were changed in Excel to reflect a date in each column before being loaded into R.


```{r dataload, message=FALSE,error=FALSE, cache=TRUE}
library(readr)
library(knitr)

Crime_by_category <- read_csv("~/Dropbox/RStudio/CalgaryCrime/Data/2017 Community Crime Statistics by category.csv")
kable(head(Crime_by_category))
```

## Data conditioning

Replace NA values with zeroes and removed unused rows (Sept 2017 - December 2017).

```{r NAreplacement, message=FALSE}
library(knitr)
library(dplyr)
Crime_by_category[is.na(Crime_by_category)] <- 0
Crime_by_category <- select(Crime_by_category, -SEP:-DEC)
kable(head(Crime_by_category))
```

## Exploratory data analysis

Plot total crime stats by category. First, add the total crime stats over the 2012-2017 time period.

```{r Plotcat, fig.width=12, message=FALSE}
library(ggplot2)
Crime_by_category$total <- rowSums(Crime_by_category[,3:70])
plot <- ggplot(Crime_by_category, aes(x=Category,y=total)) +
        geom_bar(stat="identity") +
        theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
plot
```

Next plot total crime stats by community.

```{r Plotcom, message=FALSE, fig.width=12}
Crime_by_category$total <- rowSums(Crime_by_category[,3:70])
plot <- ggplot(Crime_by_category, aes(x=Community,y=total, fill=Category)) +
        geom_bar(stat="identity") +
        theme(axis.text.x = element_text(angle=45, hjust=1, vjust=.5, size = 4))
plot
```

Only plot total crime stats by community over 5000 incidents.

```{r Plotcom2, message=FALSE, fig.width=12}
library(knitr)
library(ggplot2)
Crime_by_category_high <- subset(Crime_by_category, total >2500)
plot <- ggplot(Crime_by_category_high, aes(x=Community,y=total, fill=Category)) +
        geom_bar(stat="identity") +
        theme(axis.text.x = element_text(angle=45, hjust=1, vjust=.5, size = 6))
plot

# kable(Crime_by_category_high)
```

## To do

1. Recast the data to a more suitable format, according to the tidy data methodology (http://vita.had.co.nz/papers/tidy-data.html)
2. Load Census data to normalize absolute crime rates : http://www.calgary.ca/CA/city-clerks/Pages/Election-and-information-services/Civic-Census/2017-Results.aspx
3. Georeference data to a map of the city of Calgary by integrating the work of Chel Hee Lee, as shown at the September 27th Calgary R User Group (CRUG) meeting. (http://people.ucalgary.ca/~chelhee.lee/pages/crug.html)