---
title: "Calgary Crime Analysis"
author: "Marc Boulet & Calgary R User Group"
date: "2017-10-11"
output: 
  html_document: 
    theme: spacelab
---

```{r setup, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calgary Crime Statistics - Summary

We will perform a data analysis on the Calgary Crime Statistics dataset, retrieved from the following website: http://www.calgary.ca/cps/Documents/statistical-reports/2017%20Community%20Crime%20Statistics.xls.


A few changes were made in Excel before loading into R:  
1. date column names were changed in Excel (collapsed two cells into one)  
2. blank lines were removed  

```{r dataload, message=FALSE,error=FALSE, cache=TRUE}
library(readr) # needed to load .csv files
CalgaryCrime <- read_csv("~/Dropbox/RStudio/CalgaryCrime/CalgaryCrime/2017 Community Crime Statistics by category.csv") # change URL to point to data
```

## Data conditioning

Before exploring the data, certain data conditioning steps were taken:  
1. Replace NA values with zeroes  
2. removed unused columns (Sept 2017 - December 2017).  

```{r NAreplacement, message=FALSE}
library(knitr) # used to display data using kable
library(dplyr) # used to wrangle data
CalgaryCrime[is.na(CalgaryCrime)] <- 0 # add zeroes to NA cells
CalgaryCrime <- select(CalgaryCrime, -SEP:-DEC) # remove unused columns
kable(head(CalgaryCrime)) # display resulting data
```


### Data wrangling using the Tidy Data methodology

In a tidy data set:  
1. Each **variable** is in a column.  
2. Each **observation** is a row.  
3. Each **value** is a cell.  

*Arranging your data in this way makes it easier to work with because you have a consistent way of referring to variables (as column names) and observations (as row indices). When using tidy data and tidy tools, you spend less time worrying about how to feed the output from one function into the input of another, and more time answering your questions about the data.*

```{r tidydata, message=FALSE}
library(tidyr)
CalgaryCrimeTidy <- CalgaryCrime %>% gather(Date, Cases, 3:70) # move data columns into one column
kable(head(CalgaryCrimeTidy)) # display resulting data

```

### Add Calgary census data

In order to perform more meaningful data analysis, the 2017 Civic Census Results, consisting of community populations from 2013-2017, were joined to the existing dataset. The raw data is located here: http://www.calgary.ca/CA/city-clerks/Documents/Election-and-information-services/Census2017/2017_Census_Results_Community_Tables.xlsx. 

```{r censusdata, message=FALSE,error=FALSE, cache=TRUE}
library(dplyr)
library(readr)
library(knitr)

CalgaryCensus <- read_csv("~/Dropbox/RStudio/CalgaryCrime/CalgaryCrime/2017_Census_Results_Community_Population.csv")
CalgaryCensus$AvgPop <- rowMeans(subset(CalgaryCensus, select = c(2:6))) # calc 5 yr pop average 
CalgaryData <- left_join(CalgaryCrimeTidy, CalgaryCensus, by = "Community") # add pop data to crime data
kable(head(CalgaryData)) # display resulting data
```

## Exploratory data analysis

We will generate a few plots to gain insight into the data.  

####Plot total crime stats by category  

```{r plottotalcategories, fig.width=12}
library(knitr)
library(ggplot2)
CatTotal <-   CalgaryData %>% 
                group_by(Category) %>% 
                summarise(TotalByCategory = sum(Cases)) %>%
                arrange(desc(TotalByCategory))

plot <- ggplot(CatTotal, aes(x=Category,y=TotalByCategory)) +
        geom_bar(stat="identity", fill="blue") +
        theme(axis.text.x = element_text(angle=45, hjust=1, size=10))
plot
```

#### Plot total crime stats by community 

```{r plottotalcases, fig.width=12, fig.height=8}
library(knitr)
library(ggplot2)
CasesTotal <-   CalgaryData %>% 
                group_by(Community) %>% 
                summarise(TotalByCommunity = sum(Cases)) %>%
                arrange(desc(TotalByCommunity))

plot <- ggplot(CasesTotal[1:25,], aes(x=reorder(Community, TotalByCommunity), y=TotalByCommunity)) +
        geom_bar(stat="identity", fill="red") +
        theme(axis.text.y = element_text(size=12)) +
        geom_text(aes(label=TotalByCommunity), hjust=1.2, size=5, colour="white") +
        coord_flip()
plot
```

#### Normalize total crime stats by population  

Next, we'll normalize the data according to average population in each community:

```{r normalize, fig.width=12, message=FALSE}

CasesTotal <- left_join(CasesTotal, CalgaryCensus, by = "Community")
CasesTotal$Per100 <- CasesTotal$TotalByCommunity / CasesTotal$AvgPop * 100
kable(head(CasesTotal, 15))
```

#### Outliers

However, there are a number of outliers that occur:  
1. data with NA values (missing population data)  
2. data with zero values (zero population data)    
3. data with exceptionally high normalized values (due to high crime rate & low average population, eg. industrial parks)  

```{r outliers}
Outliers <- CasesTotal %>% 
            filter(!complete.cases(Per100) | Per100 =="Inf" | Per100 > 500)
kable(head(Outliers, 10))
```

#### Plot total crime stats by population (removing outliers)

```{r plotnormalisedtotals, fig.width=12,fig.height=8, message=FALSE}
library(ggplot2)

CasesTotalClean <-  CasesTotal %>% 
                    filter(complete.cases(Per100) & Per100 != "Inf" & Per100 < 500 & AvgPop > 500) %>%
                    arrange(desc(Per100))

plot <- ggplot(CasesTotalClean[1:25,], aes(x=reorder(Community, Per100), y=Per100)) +
        geom_bar(stat="identity", fill = "darkblue") +
        theme(axis.text.y = element_text(size=12)) +
        geom_text(aes(label= round(Per100)), hjust=1.2, size=5, colour="white") +
        geom_text(aes(label=round(AvgPop)), hjust=-0.2) +
        coord_flip()
plot
kable(head(CasesTotalClean,25))
```

## To do
1. Recast the data to a more analysis-friendly format, according to the tidy data methodology (http://vita.had.co.nz/papers/tidy-data.html) **DONE**  
2. Load Census data to normalize absolute crime rates: http://www.calgary.ca/CA/city-clerks/Pages/Election-and-information-services/Civic-Census/2017-Results.aspx **DONE**   
3. Georeference data to a map of the city of Calgary by integrating the work of Chel Hee Lee, as shown at the September 27th Calgary R User Group (CRUG) meeting. (http://people.ucalgary.ca/~chelhee.lee/pages/crug.html) 
4. Interactive comparisons between communities, categories, years, etc.